SmartCheck

### 2012-12-04 paper/paper.tex
paper updates.

\documentclass[9pt]{sigplanconf}
\usepackage{url}
\usepackage{tikz}
\usepackage{graphicx}
\usepackage{framed}
\usepackage{color}
\usepackage{url}
\usepackage{alltt}
\usepackage{tabularx}
\usepackage{subfigure}
\usepackage{fancyhdr}
\usepackage{paralist}

@@

\usepackage{listings}
\lstset{ %
  language=Haskell,                 % the language of the code
  basicstyle=\scriptsize,           % the size of the fonts that are used
  %% for the code
  %%firstnumber=0, % index to start numbering at
  %%numbers=left,                      % where to put the line-numbers
  %% numberstyle=\tiny\color{gray},  % the style that is used for the
  %% line-numbers
  stepnumber=5,                   % the step between two line-numbers. If it's
                                  % 1, each line 
  %%                              % will be numbered
  numbersep=5pt,                  % how far the line-numbers are from the code
  xleftmargin=0ex,
  %% backgroundcolor=\color{white},      % choose the background color. You must
  %% add \usepackage{color}
  %% showspaces=false,               % show spaces adding particular underscores
  %% showstringspaces=false,         % underline spaces within strings
  %% showtabs=false,                 % show tabs within strings adding
  %% particular underscores
  %% frame=single,                   % adds a frame around the code
  %% rulecolor=\color{black},        % if not set, the frame-color may be
  %% changed on line-breaks within not-black text (e.g. commens (green here))
  %% tabsize=2,                      % sets default tabsize to 2 spaces
  %% captionpos=b,                   % sets the caption-position to bottom
  %% breaklines=true,                % sets automatic line breaking
  %% breakatwhitespace=false,        % sets if automatic breaks should only
  %% happen at whitespace
  %% title=\lstname,                   % show the filename of files included
  %% with \lstinputlisting;
  %%                                 % also try caption instead of title
  %% keywordstyle=\color{blue},          % keyword style
  %% commentstyle=\color{dkgreen},       % comment style
  %% stringstyle=\color{mauve},          % string literal style
  escapeinside={@*}{*@},            % if you want to add a comment within your
                                    % code
  %% morekeywords={*,...}           % if you want to add more keywords to
  %% the set
}

@@

\newenvironment{code}{\begin{alltt}\scriptsize}{\end{alltt}}
%% \newenvironment{cols}{\begin{tabular}{m{3.6cm}|m{3.6cm}}{Haskell} &
%%     {\scriptsize Copilot}\\\hline}{\end{tabular}}

@@

instance Arbitrary B where
  arbitrary = liftM4 B arbitrary arbitrary arbitrary arbitrary
  shrink (B a b c d) = 
    [ B w x y z | w <- shrink a, x <- shrink b
                , y <- shrink c, z <- shrink d ]

@@

Typically, a user tries a brute-force approach to reduce the number of potential
counterexamples, for example, by truncating lists using the \ttp{take n} function
that returns the first $n$ elements of a list.  The trade-off is quicker
shrinking with a lower probability of finding a smaller counterexample.  For
example, redefining shrink for type \ttp{B} as follows
%
\lstset{ caption={Truncated shrinking function for data type \ttp{B}.}
       , label=lst:newshrink
       ,captionpos=b }
\begin{lstlisting}
len = 10

shrink (B a b c d) = [ B w x y z | w <- tk a, x <- tk b
                                 , y <- tk c, z <- tk d ]
  where tk x = take len (shrink x)
\end{lstlisting}
%% \begin{code}
%% \end{code}
%
\noindent
controls the exponential blowup of the state-space by taking the first ten
elements of each list, so the cross product is limited to 1000 elements.  The
downside is that potentially smaller counterexamples may be omitted.

@@

\begin{figure}[ht]
\scriptsize
  \begin{center}
    \begin{tabular}{|r||c|c|c|c|}
\hline 
 & QC (none) & QC (10) & QC (20) & SmartCheck \\      
\hline \hline 
Max.  & 0.15s & 21.51s & 125.37s & 1.96s\\
\hline
Mean  & 0.07s & 1.21s & 3.80s & 0.30s\\
\hline
Median & 0.07s & 0.48s & 0.52s & 0.24s\\
\hline
Mean size & 70 & 34 & 32 & 5\\
\hline
    \end{tabular}
  \end{center}
  \caption{Results for 100 tests of the program in Figure~\ref{fig:initial}
    using interpreted Haskell.}
  \label{fig:results}
\end{figure}

\paragraph{SmartCheck}
Motivated by these limitations of QuickCheck and SmallCheck, we have developed
\emph{SmartCheck}.  SmartCheck extends QuickCheck to efficiently and generically
find small counterexamples as well as generalize them as described in the
preceding paragraph.  To motivate the benefit of SmartCheck, consider the table
in Figure~\ref{fig:results}.  The table characterizes testing the program shown
in Figure~\ref{fig:initial} that we have described so far.  Four trials are
presented.  In the first three, we use QuickCheck to generate counterexamples.
The first column contains the results of using QuickCheck without shrinking to
provide a baseline.  The next two columns show the result of using QuickCheck
with definitions of \ttp{shrink} that limit the lists of fields of \ttp{B} to 10
(Listing~\ref{lst:newshrink}) and 20 elements (Listing~\ref{lst:newshrink}, with
\ttp{len = 20}), respectively.  (Recall that with the original definition for
shrink from Figure~\ref{fig:initial}, some counterexamples cause QuickCheck to
not return after hours of computation.)  In each row, we show the maximum, mean,
and median number of seconds required by QuickCheck to generate a counterexample
over 100 runs.  In the last row, we show the mean size of the final value
returned, measured by the taking the cumulative length of the lists of the
fields of \ttp{B}.  Notice that while typically counterexamples are returned in
a fraction of a second, there are outliers taking 21 seconds and 125 seconds for
the two tests.  Both tests roughly cut the size of the counterexample in half;
increasing the number of potential counterexamples can result in a significant
performance penalty for this example without much benefit.

@@

\paragraph{Smart reduction algorithm}
The algorithm we present for efficiently searching for new counterexamples can
be thought of as a recursive breadth-first traversal over the structure of the
current counterexample value, where the value is represented as a tree.  At each
node, during the traversal, we generate arbitrary \emph{structurally} smaller
sub-values and build a new value from that, leaving the remainder of the tree
unchanged.  Intuitively, by a structurally smaller value, we mean one with fewer
constructors.  We continue until we reach a fixed-point.

@@

replace tree 2 (SubVal L) = 
  B\sub{0} (B\sub{1} L\sub{3} 
        (B\sub{4} L\sub{6} L\sub{8}))
     L\sub{2}
\end{code}
%
\noindent
We will sometimes refer to unwrapped value returned by \ttp{index a i} as the
\emph{$i$th sub-value of $a$}, so for example,

\begin{figure}[ht]
  \lstset{caption={}
%         ,xleftmargin=5ex,firstnumber=0,numbers=left
         }
  \begin{lstlisting}
reduce :: SubTypes a => a -> Property a -> IO a
reduce cex prop = reduce' 1
  where
  reduce' idx =
    case index cex idx of 
      Nothing -> return cex
      Just v  -> 
        case test cex idx v prop of
          Nothing   -> reduce' (idx + 1)
          Just cex' -> reduce cex' prop

test :: SubTypes a 
     => a -> Int -> SubVal -> Property a -> Maybe a
test cex idx v prop = go maxTries
  where
  go 0 = Nothing
  go n = let v'   = newVal v in
         let cex' = replace cex idx v' in
         if prop cex' then go (n - 1) else Just cex'

  newVal (SubVal a) = sizedArbitrary a (size a)
  \end{lstlisting}
  \caption{Counterexample reduction algorithm\label{fig:reduction}}
\end{figure}

Figure~\ref{fig:reduction} shows the reduction algorithm.  The function
\ttp{reduce} takes a counterexample \ttp{cex} and the property \ttp{prop}, which
for our purposes, we can suppose has the type \ttp{a -> Bool}.  The reduction
begins at the first sub-value of \ttp{cex}.  Why not begin with \ttp{cex}
itself?  The idea is that

\todo{optimization: check value directly}

@@

\section{Generalizing Data}

@@

@@

\todo{\cite{syb} paper is motivated by generic shrink}
\todo{shrink paper rehger pointed to ``Automatic isolation of compiler errors'' from 1994}

>>>>

% can be 9pt
\documentclass[10pt]{sigplanconf}
\usepackage{url}
\usepackage{tikz}
\usepackage{graphicx}
\usepackage{framed}
\usepackage{color}
\usepackage{url}
\usepackage{alltt}
\usepackage{tabularx}
\usepackage{subfigure}
\usepackage{fancyhdr}
\usepackage{paralist}

@@

\usepackage{listings}
\lstset{ %
  language=Haskell,                 % the language of the code
  %%basicstyle= \scriptsize,           % the size of the fonts that are used
  %% for the code
  %%firstnumber=0, % index to start numbering at
  %%numbers=left,                      % where to put the line-numbers
  %% numberstyle=\tiny\color{gray},  % the style that is used for the
  %% line-numbers
  %% stepnumber=5,                   % the step between two line-numbers. If it's
                                  % 1, each line 
  %%                              % will be numbered
  %% numbersep=5pt,                  % how far the line-numbers are from the code
  %% xleftmargin=0ex,
  %% backgroundcolor=\color{white},      % choose the background color. You must
  %% add \usepackage{color}
  %% showspaces=false,               % show spaces adding particular underscores
  %% showstringspaces=false,         % underline spaces within strings
  %% showtabs=false,                 % show tabs within strings adding
  %% particular underscores
  %% frame=single,                   % adds a frame around the code
  %% rulecolor=\color{black},        % if not set, the frame-color may be
  %% changed on line-breaks within not-black text (e.g. commens (green here))
  %% tabsize=2,                      % sets default tabsize to 2 spaces
  %% captionpos=b,                   % sets the caption-position to bottom
  %% breaklines=true,                % sets automatic line breaking
  %% breakatwhitespace=false,        % sets if automatic breaks should only
  %% happen at whitespace
  %% title=\lstname,                   % show the filename of files included
  %% with \lstinputlisting;
  %%                                 % also try caption instead of title
  %% keywordstyle=\color{blue},          % keyword style
  %% commentstyle=\color{dkgreen},       % comment style
  %% stringstyle=\color{mauve},          % string literal style
  escapeinside={@*}{*@},            % if you want to add a comment within your
                                    % code
  %% morekeywords={*,...}           % if you want to add more keywords to
  %% the set
}

@@

\newenvironment{code}{\begin{alltt}}{\end{alltt}}
%% \newenvironment{cols}{\begin{tabular}{m{3.6cm}|m{3.6cm}}{Haskell} &

@@

instance Arbitrary B where
  arbitrary = liftM4 B arbitrary arbitrary 
                       arbitrary arbitrary
  shrink (B a b c d) = 
    [ B w x y z | w <- shrink a
                , x <- shrink b
                , y <- shrink c
                , z <- shrink d ]

@@

Typically, a user tries a brute-force approach to reduce the number of potential
counterexamples, for example, by truncating lists using the \ttp{take n} function
that returns the first $n$ elements of a list.  The trade-off is quicker
shrinking with a lower probability of finding a smaller counterexample.  For
example, redefining shrink for type \ttp{B} as follows
%

%% \begin{minipage}{\columnwidth}
%% \lstset{ caption={Truncated shrinking function for data type \ttp{B}.}
%%        , label=lst:newshrink
%%        , captionpos=b }
%% \begin{lstlisting}

\begin{figure}[ht]
\begin{code}
len = 10

shrink (B a b c d) = 
  [ B w x y z | w <- tk a, x <- tk b
              , y <- tk c, z <- tk d ]
  where tk x = take len (shrink x)
\end{code}
  \caption{Truncated shrinking function for data type \ttp{B}.}
  \label{lst:newshrink}
\end{figure}

%% \end{lstlisting}
%% \end{minipage}
%
\noindent
controls the exponential blowup of the state-space by taking the first ten
elements of each list, so the cross product is limited to 1000 elements.  The
downside is that potentially smaller counterexamples may be omitted.

@@

\begin{figure}[ht]
\footnotesize
  \begin{center}
    \begin{tabular}{|r||c|c|c|c|}
\hline 
 & QC (none) & QC (10) & QC (20) & SmartCheck \\      
\hline \hline 
Max.  & 0.15s & 21.51s & 125.37s & 1.96s\\
\hline
Mean  & 0.07s & 1.21s & 3.80s & 0.30s\\
\hline
Median & 0.07s & 0.48s & 0.52s & 0.24s\\
\hline
Mean size & 70 & 34 & 32 & 5\\
\hline
    \end{tabular}
  \end{center}
  \caption{Results for 100 tests of the program in Figure~\ref{fig:initial}
    using interpreted Haskell.}
  \label{fig:results}
\end{figure}

\paragraph{SmartCheck}
Motivated by these limitations of QuickCheck and SmallCheck, we have developed
\emph{SmartCheck}.  SmartCheck extends QuickCheck to efficiently and generically
find small counterexamples as well as generalize them as described in the
preceding paragraph.  To motivate the benefit of SmartCheck, consider the table
in Figure~\ref{fig:results}.  The table characterizes testing the program shown
in Figure~\ref{fig:initial} that we have described so far.  Four trials are
presented.  In the first three, we use QuickCheck to generate counterexamples.
The first column contains the results of using QuickCheck without shrinking to
provide a baseline.  The next two columns show the result of using QuickCheck
with definitions of \ttp{shrink} that limit the lists of fields of \ttp{B} to 10
(Figure~\ref{lst:newshrink}) and 20 elements (Figure~\ref{lst:newshrink}, with
\ttp{len = 20}), respectively.  (Recall that with the original definition for
shrink from Figure~\ref{fig:initial}, some counterexamples cause QuickCheck to
not return after hours of computation.)  In each row, we show the maximum, mean,
and median number of seconds required by QuickCheck to generate a counterexample
over 100 runs.  In the last row, we show the mean size of the final value
returned, measured by the taking the cumulative length of the lists of the
fields of \ttp{B}.  Notice that while typically counterexamples are returned in
a fraction of a second, there are outliers taking 21 seconds and 125 seconds for
the two tests.  Both tests roughly cut the size of the counterexample in half;
increasing the number of potential counterexamples can result in a significant
performance penalty for this example without much benefit.

@@

\subsection{Reduction Algorithm Overview}
The algorithm we present for efficiently searching for new counterexamples can
be thought of as a recursive breadth-first traversal over the structure of the
current counterexample value, where the value is represented as a tree.  At each
node, during the traversal, we generate arbitrary \emph{structurally} smaller
sub-values and build a new value from that, leaving the remainder of the tree
unchanged.  Intuitively, by a structurally smaller value, we mean one with fewer
constructors.  We continue until we reach a fixed-point.

@@

replace tree 2 (SubVal L) = 
  B\sub{0} (B\sub{1} L\sub{3} 
        (B\sub{4} L\sub{6} L\sub{8}))
     L
\end{code}
%
\noindent
We will sometimes refer to unwrapped value returned by \ttp{index a i} as the
\emph{$i$th sub-value of $a$}, so for example, \ttp{B\sub{4} L\sub{6} L\sub{8}}
is the 4th sub-value of \ttp{tree}.

\begin{figure}[ht]
  \lstset{caption={}
%         ,xleftmargin=5ex,firstnumber=0,numbers=left
         }
  \begin{lstlisting}
reduce :: SubTypes a 
       => a -> (a -> Property) -> IO a
reduce cex prop = reduce' 1
  where
  reduce' idx = 
    case index cex idx of 
      Nothing -> return cex
      Just v  -> do
        vs <- newVals v
        case test cex idx vs prop of
          Nothing   -> reduce' (idx+1)
          Just cex' -> reduce cex' prop

  newVals (SubVal a) = 
    sequence (replicate maxTries s)
    where s =     sizedArbitrary a 
              >>= return . SubVal

test :: SubTypes a 
     => a -> Int -> [SubVal] 
     -> (a -> Property) -> Maybe a
test cex idx vs prop = go vs
  where
  go []      = Nothing
  go (v:vs') = 
    let cex' = replace cex idx v in
    if pass prop cex' then go vs' 
      else Just cex'
  \end{lstlisting}
  \caption{Counterexample reduction algorithm\label{fig:reduction}}
\end{figure}

Figure~\ref{fig:reduction} shows the reduction algorithm.  The function
\ttp{reduce} takes a counterexample \ttp{cex} and the property \ttp{prop}.  At
the beginning of execution, we assume that a counterexample has been obtained
somehow; our implementation relies on QuickCheck as a ``front-end'' to generate
a counterexample that the algorithm automatically shrinks.  The reduction begins
at the first sub-value of \ttp{cex}, \ttp{v}.  When the index \ttp{idx} becomes
out-of-bounds and returns \ttp{Nothing}, the algorithm terminates.  Otherwise, a
list of new random values are generated.
%
\begin{code}
sizedArbitrary :: SubTypes a => a -> IO a  
\end{code}
%
generates a new value \ttp{v'} having the same type as \ttp{v} and strictly
smaller (with respect to the \ttp{size} method) than \ttp{v}.
\ttp{sizedArbitrary} returns a value inside the \ttp{IO} monad since it depends
on a random seed for generation (we could pass in the generator seed, but for
simplicity, we assume one is generated locally).  The constant \ttp{maxTries} is
the maximum number of tries to discover a new counterexample by replacing
\ttp{v} in \ttp{cex} and testing it.  The result of \ttp{pass prop cex'} for
%
\begin{code}
pass :: (a -> Property) -> a -> Bool
\end{code}
%
holds if \ttp{cex'} satisfies the property \ttp{prop}.  If no failure is found,
we move to the next sub-value of \ttp{cex} and continue.  However, if a new
smaller counterexample \ttp{cex'} is found, we start a new breadth-first
traversal of \ttp{cex'}, attempting to shrink it further.

\subsection{Reduction Algorithm Details and Optimizations}
The reduction algorithm description above omits some minor details and
optimizations we describe here.

\paragraph{Sum-type hypothesis}
First, recall that the algorithm begins at the \emph{first} sub-value of the
counterexample rather than the 0th sub-value.  The 0th sub-value of \ttp{cex} is
\ttp{cex} itself.  No invariant of the algorithm would be violated by beginning
with the 0th sub-value; recall that it would generate strictly smaller values
than \ttp{cex}.  However, we begin with a strict sub-value because of a
hypothesis we refer to as the \emph{sum-type hypothesis}.  The hypothesis is
that in the search space of possible values, if a counterexample of the form
\ttp{C}~$\vec{x}$, where \ttp{C} is some constructor and $\vec{x}$ are its
arguments, is discovered, then it is more likely to find another ``nearby''
counterexample of the form \ttp{C}~$\vec{y}$, where $\vec{y}$ are different
arguments to \ttp{C}.  For example, consider a property about unintended
variable capture over a language's parse tree represented by a sum type with
constructors for module imports, function definitions, and global-variable
assignments, respectively.  A function definition counterexample can only be
reduced to smaller function definition counterexamples, the only construct in
which variable capture is possible.

Another way to motivate the hypothesis is that searching the space of possible
counterexamples by generating \emph{any} well-typed value that is strictly
smaller than the original counterexample is something we have seen before: it is
precisely the naive approach to implementing \ttp{shrink}!

\paragraph{Conditional properties}
A QuickCheck property may be a conditional of the form
%
\begin{code}
prop a = pre a ==> prop' a
\end{code}
%
\noindent
such that \ttp{prop} is true if the precondition \ttp{pre a} is false or
\ttp{prop' a} holds.  In practice, conditionals are used to restrict the set of
values tested; values that fail the precondition are discarded.  Thus, the call
to \ttp{pass} in the reduction algorithm only considers a potential
counterexample \ttp{cex} to have failed if it passes the property's
precondition and fails the consequent.

\paragraph{Counterexample search}
We make two optimizations in searching for counterexamples.  First, just like
QuickCheck's \ttp{arbitrary} method does, our \ttp{sizedArbitrary} function
successfully larger counterexamples when generating new values with which to
replace a sub-value.

Furthermore, we can instrument a low-cost shortcut check in the algorithm.
Sometimes, a counterexample fails a property due to a sub-value nested deep
inside the counterexample.  For example, consider a simple calculator language
containing constant \ttp{Int}s, addition, and division:
%
\begin{code}
data Exp = C Int
         | Add Exp Exp
         | Div Exp Exp  
\end{code}
%
\noindent
and consider an evaluator for the language that returns \ttp{Nothing} if
division-by-zero is encountered and \ttp{Just} the value otherwise:
%
\begin{code}
eval :: Exp -> Maybe Int
eval (C i) = Just i
eval (Add e0 e1) =
  liftM2 (+) (eval e0) (eval e1)
eval (Div e0 e1) = 
  let e = eval e1 in 
  if e == Just 0 then Nothing 
    else liftM2 div (eval e0) e
\end{code}
%
\noindent
Suppose we restrict the set of programs with the following predicate:
%
\begin{code}
divSubTerms :: Exp -> Bool
divSubTerms (C _)         = True
divSubTerms (Div _ (C 0)) = False
divSubTerms (Add e0 e1)   =    divSubTerms e0 
                            && divSubTerms e1
divSubTerms (Div e0 e1)   =    divSubTerms e0 
                            && divSubTerms e1  
\end{code}
%
and then test the property
%
\begin{code}
div_prop e = divSubTerms e ==> eval e /= Nothing  
\end{code}
%
\noindent
The property \ttp{div\_prop} turns out to be false; for example, 
%
\begin{code}
Div (C 1) (Div (C 0) (C 1))
\end{code}
%
\noindent
is a counterexample.  In the predicate case
%
\begin{code}
divSubTerms (Div _ (C 0)) = False  
\end{code}
%
\noindent
we forgot to recur on the divisor.  However, a typical counterexample returned
by QuickCheck, without \ttp{shrink} implemented, is as follows:
%
\begin{quote}
\ttp{Div (Div (C (-22)) (Div (C 2) (C 13))) (Div (Div (Div (C (-11)) (C 14))
  (Add (Div (C (-4)) (C (-3))) (Div (C (-25)) (C (-25))))) (Add (Add (C 22) (Div
  (C (-30)) (C (-14)))) (C 13)))}
\end{quote}
%
\noindent
To butcher an expression, finding the cause of failure is like searching for a
divide-by-zero in the haystack!  But it turns out that the culprit is a small
sub-value:
%
\begin{code}
Div (C (-22)) (Div (C 2) (C 13))  
\end{code}
%

Generalizing this scenario, during the reduction algorithm's breadth-first
search through a counterexample \ttp{cex}'s sub-values, we may happen upon a
sub-value \ttp{cex'} that has the same type as \ttp{cex} and fails the property
(while passing any preconditions).  In this case, we can return \ttp{cex'}
directly, and rerun the reduction algorithm on \ttp{cex'}.

\todo{violates sum-type hypothesis but cheap---not testing generated values,
  just direct subvalues.  Call this the sub-value hypothesis: direct sub-values
  of a counterexample may also be a counterexample.}

\paragraph{Bounding counterexample exploration}
\todo{limit depth}
\todo{limit size of generated structures}

\paragraph{Base types}

@@

\section{Counterexample Generalization}

@@

\todo{generics}

@@

\todo{shrink paper rehger pointed to ``Automatic isolation of compiler errors'' from 1994}
